---
title: "clusterProfiler-KEGG-analysis"
author: "Eileen Owens"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = TRUE,
                      cache = TRUE,
                      warning = FALSE)
```

# Install necessary software packages (first time only):
```{r installation, eval=FALSE, results="hide", message=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#         install.packages("BiocManager")
BiocManager::install(version = "3.14") # This version of Bioconductor works with R version 4.1. Use "3.11" for R version 4.0
BiocManager::install("clusterProfiler")
BiocManager::install("pathview")
BiocManager::install("enrichplot")
install.packages("ggnewscale")
```

## Load necessary software packages (every time):
```{r software-load, results="hide", message=FALSE}
library(clusterProfiler)
library(pathview)
library(enrichplot)
library(ggplot2)
library(ggnewscale)
library(stringr)
```

# Prepare input
## Import DESeq2 data
Input: DESeq2 results in .CSV format for various comparisons of interest.
```{r}
# Set working directory:
setwd("/path/to/02_scripts") # set working directory to local scripts directory
# Read in data from deseq2
df1 = read.csv("../03_output/DESeq2res.csv", header=TRUE) # replace with correct path and filename of DESeq2 results csv file

geneList <- df1$stat # Creates a vector of the stat column from DESeq2 output
names(geneList) <- df1$gene_name # Names vector with gene names
```

# KEGG Gene Set Enrichment Analysis

## Convert Ensembl IDs to Entrez Gene IDs
toType in the bitr function below has to be one of the available options from keyTypes(org.Cf.eg.db) and must map to one of "kegg", "ncbi-geneid", "ncbi-proteinid" or "uniprot" because gseKEGG() only accpets one of these 4 options as its keytype parameter.

```{r}

## Install and load your reference genome annotations. All available annotations can be found here: http://bioconductor.org/packages/release/BiocViews.html#___OrgDb.
organism = "org.Cf.eg.db" # Canis lupus familiaris
#BiocManager::install(organism, character.only=TRUE) # First time only, then comment out this line.
library(organism, character.only=TRUE)

# See which keytypes are available for your reference genome. "SYMBOL" and "ENTREZID" are required for this script to work as written.
keytypes(org.Cf.eg.db)
```

Convert gene IDs for gseKEGG function. We will lose some genes here because not all IDs will be converted.
```{r}
ids <- bitr(names(geneList), fromType = "SYMBOL", toType ="ENTREZID", OrgDb=organism) # It is normal for this call to produce some messages/warnings.

# Remove duplicate IDS
dedup_ids <- ids[!duplicated(ids[c("SYMBOL")]),] # adjust keytype as necessary

# Create a new dataframe, df2, which has only the genes which were successfully mapped using the bitr function.
df2 <- df1[df1$gene_name %in% dedup_ids$SYMBOL,]
df2 <- df2[!duplicated(df2[c("gene_name")]),] # I've found that sometimes this extra step is needed to remove all duplicates

# Create a new column in df2 with the corresponding ENTREZ IDs
df2$entrez_id <- dedup_ids$ENTREZID

# Create a vector of log2 fold change
kegg_gene_list <- df2$log2FoldChange

# Name vector with ENTREZ ids
names(kegg_gene_list) <- df2$entrez_id

# Omit any NA values
kegg_gene_list <- na.omit(kegg_gene_list)

# Filter for only genes with log2fc >2:
kegg_mostup <- names(kegg_gene_list)[abs(kegg_gene_list) > 2]
df3 <- df2[df2$entrez_id %in% kegg_mostup,]
kegg_mostup <- df3$log2FoldChange
names(kegg_mostup) <- df3$entrez_id

# Sort both gene lists in decreasing order
kegg_gene_list = sort(kegg_gene_list, decreasing = TRUE)
kegg_mostup = sort(kegg_mostup, decreasing = TRUE)
```

## Create gseKEGG object
KEGG organism codes can be found here: https://www.genome.jp/kegg/catalog/org_list.html
```{r, warning=FALSE}
kegg_organism = "cfa" # KEGG organism code for Canis lupus familiaris
kk <- gseKEGG(geneList = kegg_mostup, # only genes with log2fc > 2 included here; can use kegg_gene_list for ranked list of all genes if desired
               organism = kegg_organism,
               nPerm = 10000, # the higher the number of permutations, the more accurate your result
               minGSSize = 3,
               maxGSSize = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType = "ncbi-geneid")
```

# Visualization of KEGG Enrichment Analysis Results

## Dotplot

### Version 1
```{r}
dotplot(kk, showCategory=10, title = "Enriched Pathways", split=".sign") + facet_grid(.~.sign)
```

### Version 2
```{r, fig.height=8, fig.width=10}
kk %>%
  dotplot(showCategory = 30, x = "NES") +
  scale_colour_viridis_c(name = "Adjusted\nP-value",
                         option="H") +
  scale_y_discrete(labels = ~ str_wrap(gsub('_', ' ', .x), 40)) + # replaces underscores with spaces to allow wrapping of long gene set names
  geom_vline(xintercept = 0, linetype=2) +
  labs(title="KEGG Enrichment Analysis", y = "Gene Set") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Enrichment Map
```{r, fig.height=8}
kk_2 <- pairwise_termsim(kk)
emapplot(kk_2, showCategory = 10)
```

## Category Netplot
```{r, fig.width=10}
# categorySize can be either 'pvalue' or 'geneNum'
cnetplot(kk, categorySize="pvalue", foldChange=kegg_gene_list, showCategory=5, node_label='category')
```

## Ridgeplot
```{r, fig.height=10}
ridgeplot(kk, showCategory=30) + labs(x = "enrichment distribution", title="KEGG Enrichment Analysis", y = "Gene Set") +
  scale_y_discrete(labels = ~ str_wrap(gsub('_', ' ', .x), 50)) + # replaces underscores with spaces to allow wrapping of long gene set names
  theme(plot.title = element_text(hjust = 0.5))
```


# Citations
```{r}
sessionInfo()
citation()
```
