---
title: "Gene Set Variation Analysis"
author: "Eileen Owens"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = TRUE,
                      warning = FALSE)
```


```{r load-packages, include=FALSE}
library(BiocManager)
library(DESeq2)
library(clusterProfiler)
library(corrplot)
library(RColorBrewer)
library(pheatmap)
library(dplyr)
library(readr)
library(knitr)
library(ggplot2)
library(cowplot)
library(gridExtra)
library(rtracklayer)
library(Rsamtools)
library(grid)
library(GGally)
# library(edgeR)
library(stringr)
library(matrixStats)
library(gridExtra)
library(magrittr)
library(GSVA)
library(dplyr)
library(GSEABase)
library(shinyjs)
library(shinybusy)
library(qusage)
library(msigdbr)
library(Biobase)
```

# Data Import and Preparation

## DDS object
This code requires the dds object generated with DESeq2 in the RNAseq-Analysis pipeline. It is recommended to use normalized counts for GSVA, and most individuals seem to use vst transformed DESeq object. We also need to convert probe IDs to gene symbols for analysis, and a metadata file for annotating figures.
```{r data-input}
setwd("/Users/eileen/PTCL/PROJ01_26PTCLs/02_scripts") # set working directory before importing files

## Perform vst transformation
dds <- readRDS("dds_CD4sOnly_noCF21.RData")
vsd <- assay(vst(dds))

## Convert probe IDs to gene symbols
genenames <- read.csv("../01_input/gene.description.csv") # Import metadata file of gene symbols corresponding to probe IDs.
vsd <- as.data.frame(vsd) # vsd is matrix array by default; convert to data frame for easier manipulation.
vsd$probe_id <- rownames(vsd) # make a column of probe IDs
vsd <- merge(as(vsd,"data.frame"), genenames, by="probe_id", all.x=TRUE) # Adds a column of gene names for the associated probe ID
vsd <- vsd[!duplicated(vsd$gene_name), ] # Remove rows with duplicate gene symbols.
rownames(vsd) <- vsd$gene_name # Make gene symbols the rownames
vsd <- dplyr::select(vsd, -c("probe_id", "gene_name", "description")) # Remove the probe_id, gene_name, and description columns.
head(vsd) #make sure columns are sample names and rows are gene symbols.
vsd_matrix <- data.matrix(vsd) # Convert to a matrix prior to analysis.

# Import GMT file
gmt <- getGmt("../GSEA/Input/scdata_gmt.gmt", geneIdType = SymbolIdentifier())

## Import metadata
metadata <- read.csv("../01_input/metadata.csv")
rownames(metadata) <- metadata$Sample # Make the sample ID column the rownames; change to reflect the name of the column with sample IDs in your metadata file 
metadata <- dplyr::select(metadata, c("Phenotype")) # Select the columns in your metadata file you want annotated on the GSVA heatmap
head(metadata) # Make sure samples are rows and classifiers are columns
```

## Optional: Subset samples from vst matrix
```{r subsetVST, eval=FALSE}
# Set eval=TRUE if including this chunk
## Extract list of samples beloninging to a particular group using metadata
# Define groups to keep 
keepGroups <- c("Phenotype1", "Phenotype2")
# Keep only rows and columns with samples in these groups
metadataSubset <- metadata %>%
  filter(Phenotype %in% keepGroups)
# Extract the list of sample IDs remaining
keepList <- metadataSubset$Sample

## Subset vst matrix to include only the samples in this list
vsd_matrix <- vsd_matrix[,keepList] 
```

# GSVA Analysis
```{r GSVA, message=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
# set parameters for GSVA
gsva_param <- gsvaParam(
  vsd_matrix,
  gmt,
  kcdf = "Gaussian", # Compute Gaussian-distributed scores
  minSize = 10, # Minimum gene set size
  maxSize = 500, # Maximum gene set size
  maxDiff = TRUE,
  absRanking = FALSE
)

# perform GSVA
gsva_results <- gsva(
  gsva_param,
  verbose=FALSE,   # Don't print out the progress bar
)

gsva_heatmap <- pheatmap::pheatmap(gsva_results,
                                      annotation_col = metadata,
                                      show_colnames = TRUE,
                                      show_rownames = TRUE,
                                      fontsize_row = 10,
                                      cluster_cols = TRUE, 
                                      cluster_rows = TRUE,
                                      cutree_rows = 5,
                                      cutree_cols = 3,
                                      main = "GSVA. Input: Vst transformed normalized DESeq2 counts, Clustering: Ward, Distance: Euclidean",
                                      clustering_distance_rows = "euclidean",
                                      clustering_distance_cols = "euclidean",
                                      clustering_method = "ward.D2") + 
 scale_fill_gradient(c('dodgerblue', 'black', "yellow"))# Shrink the pathway labels a tad
```


# Export results
```{r file-export}
# Cut column dendrogram into groups for sample-to-cluster assignment and export as CSV
gsva_heatmap <- pheatmap::pheatmap(gsva_results,
                                      annotation_col = metadata,
                                      show_colnames = TRUE,
                                      show_rownames = TRUE,
                                      cluster_cols = TRUE, 
                                      cluster_rows = TRUE,
                                      clustering_distance_rows = "euclidean",
                                      clustering_distance_cols = "euclidean",
                                      clustering_method = "ward.D2")
clusterCol <- sort(cutree(gsva_heatmap$tree_col, k=3)) # Exports the main 3 clusters by columns (samples)
clusterCol <- as.data.frame(clusterCol)
write.csv(clusterCol, file="../03_output/GSVA_Clusters.csv")
```

# Citations
```{r}
sessionInfo()
citation()
```
