---
title: "GSVA_ssGSEA"
author: "Eileen Owens"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = TRUE,
                      warning = FALSE)
```


```{r load-packages, include=FALSE}
library(BiocManager)
library(DESeq2)
library(clusterProfiler)
library(corrplot)
library(RColorBrewer)
library(pheatmap)
library(dplyr)
library(readr)
library(knitr)
library(ggplot2)
library(cowplot)
library(gridExtra)
library(rtracklayer)
library(Rsamtools)
library(grid)
library(GGally)
# library(edgeR)
library(stringr)
library(matrixStats)
library(gridExtra)
library(magrittr)
library(GSVA)
library(dplyr)
library(GSEABase)
library(shinyjs)
library(shinybusy)
library(qusage)
library(msigdbr)
library(Biobase)
```

# Data Import
This code requires the dds object generated with DESeq2 in the RNAseq-Analysis pipeline. It is recommended to use normalized counts for GSVA, and most individuals seem to use vst transformed DESeq object. We also need to convert probe IDs to gene symbols for analysis, and a metadata file for annotating figures.
```{r data-input}
setwd("/Users/eileen/PTCL/PROJ01_26PTCLs/02_scripts") # set working directory before importing files

# Perform vst transformation
dds <- readRDS("dds_CD4sOnly_noCF21.RData")
vsd <- assay(vst(dds))

# Convert probe IDs to gene symbols
genenames <- read.csv("../01_input/gene.description.csv") # Import metadata file of gene symbols corresponding to probe IDs.
vsd <- as.data.frame(vsd) # vsd is matrix array by default; convert to data frame for easier manipulation.
vsd$probe_id <- rownames(vsd) # make a column of probe IDs
vsd <- merge(as(vsd,"data.frame"), genenames, by="probe_id", all.x=TRUE) # Adds a column of gene names for the associated probe ID
vsd <- vsd[!duplicated(vsd$gene_name), ] # Remove rows with duplicate gene symbols.
rownames(vsd) <- vsd$gene_name # Make gene symbols the rownames
vsd <- dplyr::select(vsd, -c("probe_id", "gene_name")) # Remove the probe_id and gene_name columns.
head(vsd) #make sure columns are sample names and rows are gene symbols.
vsd_matrix <- data.matrix(vsd) # Convert to a matrix prior to analysis.

#Import GMT file
gmt <- getGmt("../GSEA/Input/scdata_gmt.gmt", geneIdType = SymbolIdentifier())

#Import metadata
metadata <- read.csv("../01_input/metadata_CD4sOnly_outlierRemoved.csv")
rownames(metadata) <- metadata$Sample # Make the sample ID column the rownames
metadata <- metadata[, -1, drop=FALSE] # Drop the sample ID column; drop=FALSE will keep rownames with only one column instead of returning a character vector.
head(metadata) # Make sure samples are rows and classifiers are columns
```

# GSVA Analysis
```{r GSVA, message=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
gsva_results <- gsva(
  vsd_matrix, # count matrix
  gmt,
  method = "gsva",
  # Appropriate for our vst transformed data
  kcdf = "Gaussian",
  # Minimum gene set size
  min.sz = 10,
  # Maximum gene set size
  max.sz = 500,
  # Compute Gaussian-distributed scores
  mx.diff = TRUE,
  # Don't print out the progress bar
  verbose = FALSE)


pathway_heatmap <- pheatmap::pheatmap(gsva_results,
                                      annotation_col = metadata,
                                      show_colnames = TRUE,
                                      show_rownames = TRUE,
                                      fontsize_row = 10,
                                      cluster_cols = TRUE, 
                                      cluster_rows = TRUE,
                                      cutree_rows = 5, 
                                      clustering_distance_rows = "euclidean",
                                      clustering_distance_cols = "euclidean",
                                      clustering_method = "ward.D2") + 
 scale_fill_gradient(c('dodgerblue', 'black', "yellow"))# Shrink the pathway labels a tad
```

# Single sample GSEA analysis
```{r ssGSEA, results=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
ssgsea_results <- gsva(
  vsd_matrix, # count matrix
  gmt, #gene sets from gmt file
  method = "ssgsea",
  # Appropriate for our vst transformed data
  kcdf = "Gaussian",
  # Minimum gene set size
  min.sz = 10,
  # Maximum gene set size
  max.sz = 500,
  # Compute Gaussian-distributed scores
  mx.diff = TRUE,
  # Don't print out the progress bar
  verbose = FALSE)

pathway_heatmap <- pheatmap::pheatmap(ssgsea_results,
                                      annotation_col = metadata,
                                      show_colnames = TRUE,
                                      show_rownames = TRUE,
                                      fontsize_row = 10,
                                      cluster_cols = TRUE, 
                                      cluster_rows = TRUE,
                                      cutree_rows = 5, 
                                      clustering_distance_rows = "euclidean", 
                                     clustering_distance_cols = "euclidean",
                                     clustering_method = "ward.D2") + 
 scale_fill_gradient(c('dodgerblue', 'black', "yellow"))# Shrink the pathway labels a tad
```

# Export results
```{r file-export}
write.csv(gsva_results, "gsva_scClusters_CD4PTCL.csv")
write.csv(ssgsea_results, "ssgsea_scClusters_CD4PTCL.csv")
```

# Citations
```{r}
sessionInfo()
citation()
```